\section{Conclusion}

In this project, we explored the Goertzel Filter as a digital signal processing algorithm for efficiently computing individual discrete Fourier transform (DFT) bins. The filter provides an alternative to the fast Fourier transform (FFT) for applications that require only a few frequency components, making it particularly useful for detecting specific frequencies in a signal. Our project focused on the design and implementation of a Goertzel Filter in VHDL, along with the development of a corresponding test bench.
The theoretical background of the Goertzel Algorithm was presented, explaining its principles and advantages. A comprehensive literature review was conducted to understand the state-of-the-art applications of the Goertzel Filter. Based on this knowledge, we proceeded with the implementation of the filter in VHDL, following the specified requirements. The implementation included a detailed description of the filter's ports, algorithm, control signals, and considerations for timing performance.
To validate the functionality and accuracy of our Goertzel Filter design, we conducted simulations using MATLAB and VHDL testbench. The MATLAB simulation with floating-point numbers allowed us to confirm our understanding of the Goertzel algorithm. Subsequently, we performed MATLAB simulations with integers and LSB truncation to verify that any deviation between MATLAB and VHDL results was due to the expected truncation error resulting from the filter design, rather than implementation mistakes in VHDL. The VHDL testbench provided further confirmation, as the results matched exactly with the MATLAB simulation, ensuring the correctness of our implementation.
In conclusion, our project successfully explored, designed, and implemented a Goertzel Filter in VHDL. Through thorough simulations and comparisons with MATLAB, we confirmed the accuracy and reliability of our implementation. The Goertzel Filter offers an efficient solution for frequency detection in various applications, and our work contributes to its understanding and practical implementation.